---
title: "Generalized Maximum Entropy and Cross-Entropy for the Linear Model"
author: "Ganbaatar Jambal"
date: "`r Sys.Date()`"
output:
  bookdown::html_document2:
    base_format: rmarkdown::html_vignette
    number_sections: true
bibliography: references.bib
csl: apa.csl
link-citations: true
vignette: >
  %\VignetteIndexEntry{Generalized Maximum Entropy and Cross-Entropy for the Linear Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
```

```{r pkg, message = FALSE}
library(infometrics)
```

# Ill-conditioned regression problems

In the companion vignette *Pure Inverse Problems* we used `inverse_ce()` to
recover a distribution from an ill-posed problem in which the number of unknowns
exceeds the number of observations. That is not the only situation in which
traditional estimation methods run into trouble. In many applied settings --- even in cases where the sample is relatively large compared to the number of parameters  --- traditional methods may still behave poorly.

Such problems are quite common in the social sciences, and especially in economics, where the data are often *observational* rather than the outcome of a carefully designed and replicable experiment. When the data are collected passively we cannot rule out possibilities of strong dependence, including in the form of multicollinearity. The consequences of the latter are well
known: standard errors are inflated, coefficients lose statistical
significance, point estimates become erratic and sensitive to small changes in
the data, and the numerical procedure becomes less stable. These situations are often encountered in impact evaluations in which the analyst adds polynomial and interaction terms. It is a sensible strategy to capture non-linearities, but it may also backfire by creating regressors with near-exact linear dependencies.

This vignette introduces the **Generalized Maximum Entropy (GME)** and
**Generalized Cross-Entropy (GCE)** estimators [@golan1996maximum; @golan2008information], which are designed to deal with these ill-conditioned problems, and shows how to fit them with the `linreg()` function of the infometrics package. At the end of this vignette, we also demonstrate its sibling, the `linreg_iv()` function, which deals with instrumental variables.

But before demonstrating a working example, let's highlight the underlying concepts and core mechanics of the GME and GCE estimators, and start with the general linear model (GLM)

\begin{equation}
  \mathbf{y} = \mathbf{X}\boldsymbol{\beta} + \mathbf{e}
  (\#eq:glm)
\end{equation}

where $\mathbf{y} = (y_1, \dots, y_T)'$ is the $T$-dimensional response,
$\mathbf{X}$ is a known $T \times K$ matrix of explanatory variables,
$\boldsymbol{\beta} = (\beta_1, \dots, \beta_K)'$ is the $K$-vector of unknown
parameters, and $\mathbf{e} = (e_1, \dots, e_T)'$ is the noise, which absorbs
sampling error, non-sampling error, and model misspecification.

# The GME formulation

The idea behind GME is to treat the unknown coefficients and errors as *random
variables* and to reparameterize them as expected values over bounded **support
spaces**. Each estimated coefficient is written as the mean of a discrete distribution
$\mathbf{p}_k$ on a support $\mathbf{z}_k = (z_{k1}, \dots, z_{kM})$, and each
error as the mean of a distribution $\mathbf{w}_t$ on a support
$\mathbf{v} = (v_1, \dots, v_J)$:

\begin{equation}
  \beta_k = \sum_{m=1}^M z_{km}\, p_{km},
  \qquad
  e_t = \sum_{j=1}^J v_j\, w_{tj}.
  (\#eq:reparam)
\end{equation}

Because $\mathbf{P}$ and $\mathbf{W}$ are probability distributions, they must
satisfy the additivity (normalization) constraints

\begin{equation}
  \sum_{m=1}^M p_{km} = 1 \;\; (k = 1, \dots, K), \qquad
  \sum_{j=1}^J w_{tj} = 1 \;\; (t = 1, \dots, T),
  (\#eq:add)
\end{equation}

with $p_{km} \ge 0$ and $w_{tj} \ge 0$. The support $\mathbf{z}_k$ must be wide
enough to contain the true value of $\beta_k$; the error support $\mathbf{v}$ is
symmetric around zero so that the noise is mean-zero, with a half-width set by
the **three-sigma rule** [@pukelsheim1994three] --- three times an estimate of
$\mathbf{y}$'s standard deviation.

Among the infinitely many distributions consistent with the data we select the
one that is maximally uninformative, i.e. that maximizes the Shannon entropy
[@shannon1948mathematical]

\begin{equation}
  H(\mathbf{P}, \mathbf{W}) =
    - \sum_{k}\sum_{m} p_{km} \log p_{km}
    - \sum_{t}\sum_{j} w_{tj} \log w_{tj}.
  (\#eq:entropy)
\end{equation}

Maximizing \@ref(eq:entropy) subject to the data constraint \@ref(eq:glm), the reparameterization \@ref(eq:reparam) and the normalizations \@ref(eq:add) yields the following solutions [see @golan1996maximum, Ch. 6; @golan2008information, Sec. 6.1]

\begin{equation}
  \hat{p}_{km}
    = \frac{\exp\!\left(-z_{km} \sum_t \hat{\lambda}_t x_{tk}\right)}
           {\Omega_k(\hat{\boldsymbol{\lambda}})},
  \qquad
  \hat{w}_{tj}
    = \frac{\exp\!\left(-\hat{\lambda}_t v_{j}\right)}
           {\Psi_t(\hat{\boldsymbol{\lambda}})},
  (\#eq:solution)
\end{equation}

where $\Omega_k$ and $\Psi_t$ are the normalizing (partition) functions and
$\hat{\boldsymbol{\lambda}}$ is a $T$-vector of Lagrange multipliers on the data
constraint. As with the pure inverse problem, $\lambda_t$ measures the
information content of observation $t$: a multiplier near zero means that row
carries little information.

# From the primal to the concentrated dual

Substituting the solution \@ref(eq:solution) back into the Lagrangian collapses
the primal problem into an unconstrained **concentrated (dual)** problem in
$\boldsymbol{\lambda}$ alone:

\begin{equation}
  \min_{\boldsymbol{\lambda}}
    \left\{
      \sum_t y_t \lambda_t
      + \sum_k \log \Omega_k(\boldsymbol{\lambda})
      + \sum_t \log \Psi_t(\boldsymbol{\lambda})
    \right\}.
  (\#eq:dual)
\end{equation}

Please note that the $\boldsymbol{\lambda}$'s are the Lagrange multipliers associated with equation \@ref(eq:glm) only; we omit the Lagrange multipliers of the normalization constraints \@ref(eq:add) because the closed-form solutions in \@ref(eq:solution) already satisfy the normalization by construction. It is also worth noting that the gradient of \@ref(eq:dual) is the moment condition
$\mathbf{y} - \mathbf{X}\boldsymbol{\beta} - \mathbf{e} = \mathbf{0}$, so at the
optimum the data are reproduced exactly, $\mathbf{y} = \mathbf{X}\hat{\boldsymbol{\beta}} + \hat{\mathbf{e}}$;
the estimator simply chooses the least-informative split of $\mathbf{y}$ into
signal and noise.

Solving in $\boldsymbol{\lambda}$ rather than in $(\mathbf{P}, \mathbf{W})$ is a
large computational saving. Suppose $T = 100$, $K = 5$, $M = 3$, and $J = 3$.
The primal problem requires estimation of $K \times M + T \times J = 15 + 300 = 315$
unknowns, whereas the concentrated dual is solved in just
$T = 100$ Lagrange multipliers. `linreg()` always uses the dual form (BFGS method with an
analytic gradient), recovering $\hat{\boldsymbol{\lambda}}$ and then mapping it
back through \@ref(eq:solution) and \@ref(eq:reparam) to recover 
$\hat{\boldsymbol{\beta}}$ and $\hat{\mathbf{e}}$.

# The GCE extension: incorporating prior information

If we have prior beliefs about the data-generating process and the noise --- summarized as prior probabilities $\left\{ \mathbf{p}^0_k, \mathbf{w}^0_t \right\}$ over their respective support spaces $\left\{ \mathbf{z}_k, \mathbf{v} \right\}$ --- we can fold them in by replacing the entropy with the **cross-entropy** (or Kullback--Leibler divergence) between the estimated distributions and their prior information:

\begin{equation}
  D(\mathbf{P} \,\Vert\, \mathbf{P}^0) + D(\mathbf{W} \,\Vert\, \mathbf{W}^0)
    = \sum_k \sum_m p_{km} \log\!\left(p_{km} / p^0_{km}\right) +
    \sum_t \sum_j w_{tj} \log\!\left(w_{tj} / w^0_{tj}\right).
  (\#eq:ce)
\end{equation}

Minimizing the divergence \@ref(eq:ce) subject to the same constraints is the GCE estimator. A **uniform** prior reproduces GME; a non-uniform prior pulls the estimates toward the prior mean. In `linreg()` prior information on $\boldsymbol{\beta}$ can be summarized in the single argument `p0` (a $K \times M$ matrix whose rows sum to one), and our subjective beliefs on $\mathbf{e}$ can be incorporated in `w0` (a $T \times J$ matrix whose rows sum to one): you may leave either one or both blank to opt for the default (uniform distribution), or supply them for GCE.

# Worked examples

We now put GME and GCE estimation to work on a deliberately ill-conditioned data set, in the spirit of the collinearity illustration in Section 6.6 of @golan2008information. `linreg()` also lets you place weights on the signal and noise, which the worked example presents as well.  

## A collinear design: GME versus OLS

We generate 200 observations from
$y = 1 - 2 x_1 + 3 x_2 + 0 \cdot x_3 + e$, where $x_1$ is a clean regressor but
$x_3$ is nearly a copy of $x_2$. The coefficient on $x_3$ is truly zero, so a
good estimator should (a) recover the strong, clean effect of $x_1$, and
(b) not be fooled by the collinear $x_2/x_3$ pair.

```{r dgp}
set.seed(10)
n  <- 200
x1 <- runif(n, 0, 20)                 # clean regressor
x2 <- runif(n, 0, 20)
x3 <- x2 + rnorm(n, 0, 0.05)          # near-collinear with x2 (extreme)

b  <- c(1, -2, 3, 0)                  # (Intercept), x1, x2, x3
y  <- b[1] + b[2] * x1 + b[3] * x2 + b[4] * x3 + rnorm(n, 0, sqrt(2))
dat <- data.frame(y, x1, x2, x3)

truth <- c(`(Intercept)` = 1, x1 = -2, x2 = 3, x3 = 0)
cor(x2, x3)                           # near-perfect collinearity
```

We fit OLS and GME on the same data. For GME we place every coefficient on a
symmetric five-point support that comfortably contains the true values, and we
set the noise support by the three-sigma rule using the standard deviation of $\mathbf{y}$.

```{r fit1}
ols <- lm(y ~ x1 + x2 + x3, data = dat)

Zvec <- seq(-6, 6, length.out = 5)   # coefficient support (shared)
sig  <- sd(dat$y)                    # standard deviation of y for the noise support
vvec <- seq(-3 * sig, 3 * sig, length.out = 5)

gme <- linreg(y ~ x1 + x2 + x3, data = dat,
              Z = Zvec, v = vvec)             # nu = 0.5, uniform priors => GME
```

```{r table1}
se <- function(fit) sqrt(diag(vcov(fit)))
tab1 <- data.frame(
  Truth    = truth,
  OLS      = coef(ols),
  `OLS SE` = se(ols),
  GME      = coef(gme),
  `GME SE` = se(gme),
  check.names = FALSE
)
knitr::kable(tab1, digits = 3,
             caption = "OLS versus GME under near-perfect collinearity of x2 and x3.")
```

The clean coefficient on $x_1$ is recovered well by both methods. The collinear
pair is the real difficulty. Because $x_2$ and $x_3$ are almost identical
($\mathrm{cor} \approx 0.99996$), the data identify only their *combined* effect
of about $3$; how OLS divides that effect between the two is essentially
arbitrary and swings wildly from sample to sample. In this draw OLS loads almost
the entire effect onto the *truly-irrelevant* $x_3$ ($\hat\beta_3 \approx 3$) and
leaves $x_2$ near zero ($\hat\beta_2 \approx 0$) --- the exact reverse of the
split a different sample would give --- while flagging the indeterminacy with
large standard errors on both. GME cannot conjure information the data do not
contain, but its bounded supports keep the estimates from exploding: it splits
the shared effect roughly evenly and stays finite and stable. Section
\@ref(adding-prior-information-gce) shows how a modest prior resolves the
attribution that the data alone cannot.

## Adding prior information (GCE)

Suppose our prior knowledge can be summarized in the following statement: "$\beta_2 \approx 2.8$", which is still biased but close to the true value. Then we can incorporate this into the GCE estimator as follows:

```{r prior}
# we can recover probabilities associated with the  
# prior information, using the inverse_ce() function  
temp <- data.frame(y=2.8, s1 = Zvec[1], s2 = Zvec[2], 
                   s3= Zvec[3], s4= Zvec[4], s5= Zvec[5])
p0_x2 <- inverse_ce(y ~ s1 + s2 + s3 + s4 + s5 - 1, data=temp)$p_hat
p0_other_x <- rep(1/5, 5)
p0 <- rbind(p0_other_x, p0_other_x, p0_x2, p0_other_x)

gce <- linreg(y ~ x1 + x2 + x3, data = dat,
              Z = Zvec, v = vvec, p0 = p0)    # GCE: informative signal prior
```

```{r table2}
tab2 <- data.frame(
  Truth          = truth,
  `OLS`      = coef(ols),
  `GME` = coef(gme),
  `GCE`   = coef(gce),
  check.names     = FALSE
)
knitr::kable(tab2, digits = 3,
             caption = "OLS, GME (uniform prior), and GCE (informative prior).")
```

With a well-chosen prior, GCE resolves the $x_2/x_3$ ambiguity that stumped OLS:
the estimate for the irrelevant $x_3$ is pulled toward zero and the effect it
had absorbed is returned to $x_2$. This is the sense in which cross-entropy is
an efficient information-processing rule [@zellner1988optimal] --- it combines
the prior with the data into a posterior-like estimate. A word of caution: a
prior is a modelling assumption, so a *wrong* prior will bias the estimates just
as effectively as a right one sharpens them.

## The effect of $\nu$ {#the-effect-of-nu}

Finally we vary the weight $\nu$, which governs how the estimator balances the
signal-entropy and noise-entropy terms in the dual \@ref(eq:dual). We refit the
GCE model (uniform priors - except for $x_2$ - same supports) at $\nu = 0.1$, $0.5$, and $0.9$ and
compare the coefficient estimates with the truth.

```{r nu}
nus <- c(0.1, 0.5, 0.9)
fit_nu <- lapply(nus, function(nu)
  linreg(y ~ x1 + x2 + x3, data = dat, Z = Zvec, v = vvec, p0=p0, nu = nu))

tab3 <- rbind(sapply(fit_nu, coef),
              `R-squared` = sapply(fit_nu, `[[`, "r.squared"),
              `Signal S`  = sapply(fit_nu, `[[`, "S"))
tab3 <- cbind(Truth = c(truth, `R-squared` = NA, `Signal S` = NA), round(tab3, 4))
colnames(tab3) <- c("Truth", paste0("nu = ", nus))
knitr::kable(tab3, digits = 3,
             caption = "Coefficients, ordinary R-squared, and signal normalized entropy across nu.")
```

The moment condition $\mathbf{y} = \mathbf{X}\hat{\boldsymbol{\beta}} +
\hat{\mathbf{e}}$ holds for every $\nu$; what changes is how much of $\mathbf{y}$
is attributed to the systematic part $\mathbf{X}\boldsymbol{\beta}$ versus the
noise $\mathbf{e}$. A larger $\nu$ shifts weight onto the noise term, a smaller
$\nu$ onto the coefficients, so the coefficients, the ordinary $R^2$, and the
signal normalized entropy $S$ (Section \@ref(inference)) all move together, as
the two summary rows show. Unless there is a specific reason to emphasize one
component over the other, the equal-weight default $\nu = 0.5$ is the
recommended choice.

# Inference: entropy diagnostics and hypothesis tests {#inference}

Section 6.4 of @golan2008information equips the GME/GCE estimator with
information-theoretic analogues of the usual regression diagnostics. All of them
are by-products of the fitted probability distributions, so they cost nothing
extra to report.

## Normalized entropy and the pseudo-$R^2$

Let's start with the normalized entropy 

\begin{equation}
  S(\hat{\mathbf{p}}) = \frac{- \sum_k \sum_m \hat{p}_{km} \log \hat{p}_{km}}{K \log M} \quad \text{(uniform priors)}, \\
  S(\hat{\mathbf{p}}) = \frac{- \sum_k \sum_m \hat{p}_{km} \log \hat{p}_{km}}{- \sum_k \sum_m p^0_{km} \log p^0_{km}} \quad \text{(otherwise)}.
  (\#eq:norm-entropy)
\end{equation}

For a uniform prior $S(\hat{\mathbf{p}})$ lies in the $[0,1]$ interval, and it is worth noting that this measure is conditional on the support chosen for the regression parameters $\boldsymbol{\beta}$. A value of $S(\hat{\mathbf{p}})$ equal to one reflects complete ignorance --- the data were essentially uninformative about the unknowns we were trying to recover --- whereas a value of zero means perfect knowledge. With an informative prior, $S(\hat{\mathbf{p}})$ is measured relative to that prior and can exceed one when the data are less informative than the prior, in which case the pseudo-$R^2$ below turns negative.

One can also construct the same measure to evaluate the relevance of the information contained in each of the variables $k = 1, 2, \dots, K$:

\begin{equation}
  S(\hat{\mathbf{p}}_k) = \frac{- \sum_m \hat{p}_{km} \log (\hat{p}_{km})}{\log M} 
  (\#eq:norment)
\end{equation}

which measures how much uncertainty about $\beta_k$ remains after conditioning on the
data. By the same token as above, a value near $1$ means the data were essentially uninformative about
$\beta_k$ (the estimate stayed close to the uniform prior); a value near $0$
means $\beta_k$ is sharply pinned down. 

It is also worth noting that $S(\hat{\mathbf{p}})$ represents the information contained in the whole system, while $S(\hat{\mathbf{p}}_k)$ reflects the information in column $k$. The model-level signal normalized entropy $S(\hat{\mathbf{p}})$ is stored on the fit as `$S`.

Using $S(\hat{\mathbf{p}})$, one can calculate the **pseudo-$R^2$** as follows,

\begin{equation}
  \text{pseudo-}R^2 = 1 - S(\hat{\mathbf{p}}),
  (\#eq:pseudoR2)
\end{equation}

which shows the share of the prior uncertainty that the data remove. Unlike the ordinary
$R^2$ (variance explained, stored as `$r.squared`), it rewards *informational*
content and remains well defined even where OLS is unstable.

```{r ent}
c(signal_S    = gme$S,
  pseudo_R2   = 1 - gme$S,
  ordinary_R2 = gme$r.squared)
```

## The entropy-ratio (ER) test

The **entropy-ratio test** is the information-theoretic counterpart of the
likelihood-ratio test. To test $H_0\!: \beta_k = 0$, the model is refit with the
$k$-th coefficient's support collapsed to zero and the statistic

\begin{equation}
  \mathrm{ER}_k = 2\left[H^*_{\text{unrestricted}} - H^*_{\text{restricted},k}\right]
  \;\sim\; \chi^2_1
  (\#eq:er)
\end{equation}

is formed, where $H^* = \sum H(\mathbf{p}) + \sum H(\mathbf{w})$ is the total
signal-plus-noise entropy [@golan2008information, Sec. 6.4]. An overall test of
$H_0\!:$ all coefficients $= 0$ collapses every support row and is referred to
$\chi^2_K$. `summary()` reports both, in place of the usual Wald $z$ table.

```{r er}
summary(gme)
```

The output shows the classic collinearity signature. The clean regressor $x_1$
is decisively significant. The *individual* ER tests for the near-identical
$x_2$ and $x_3$ need not reject $H_0\!:\beta_k = 0$ --- dropping either one lets
the other absorb the shared effect --- yet the **overall** test still detects the
joint signal. Inference about the *pair* is meaningful even when inference about
each member separately is not.

## The Wald test

The ER test refits the model; a complementary route is the **Wald test**, which
uses the asymptotic covariance $\widehat{\mathrm{Cov}}(\hat{\boldsymbol{\beta}})$
(Golan 2008, p. 96), available through `vcov()`, and needs no refitting. For $q$
linear restrictions $\mathbf{R}\boldsymbol{\beta} = \mathbf{r}$,

\begin{equation}
  W = (\mathbf{R}\hat{\boldsymbol{\beta}} - \mathbf{r})'
      \left[\mathbf{R}\,\mathbf{V}\,\mathbf{R}'\right]^{-1}
      (\mathbf{R}\hat{\boldsymbol{\beta}} - \mathbf{r})
  \;\sim\; \chi^2_q,
  \qquad \mathbf{V} = \widehat{\mathrm{Cov}}(\hat{\boldsymbol{\beta}}).
  (\#eq:wald)
\end{equation}

A compact implementation on top of `coef()` and `vcov()`:

```{r wald}
wald_test <- function(fit, R, r = NULL) {
  R <- rbind(R)                                 # allow a plain vector
  if (is.null(r)) r <- rep(0, nrow(R))
  d <- as.vector(R %*% coef(fit)) - r
  W <- as.numeric(t(d) %*% solve(R %*% vcov(fit) %*% t(R), d))
  c(W = W, df = nrow(R), p.value = pchisq(W, df = nrow(R), lower.tail = FALSE))
}

# H0: beta_x3 = 0  (single restriction; equals the squared t-ratio)
wald_test(gme, c(0, 0, 0, 1))

# H0: beta_x2 = beta_x3 = 0  (joint test of the collinear pair)
wald_test(gme, rbind(c(0, 0, 1, 0),
                     c(0, 0, 0, 1)))
```

For a single restriction the Wald statistic is just the squared $t$-ratio
$(\hat\beta_k / \mathrm{se}_k)^2$. The Wald and ER tests are asymptotically
equivalent but can disagree in finite samples --- especially under the
finite-support regularization, where they draw on different information (the
curvature of the dual for Wald, a full refit for ER). As with the ER test, the
*joint* Wald test of the collinear pair stays powerful even where the individual
tests do not; note, though, that under near-perfect collinearity the pair's
covariance block is ill-conditioned, so treat the exact magnitude of the joint
$W$ with care and read it qualitatively.

# Instrumental variables with `linreg_iv()`

When a regressor is correlated with the error --- through omitted variables,
measurement error, or simultaneity --- OLS is biased and inconsistent. The
classical remedy is an *instrument*: a variable that shifts the endogenous
regressor but is unrelated to the error. `linreg_iv()` is the
instrumental-variables sibling of `linreg()`. It identifies
$\boldsymbol{\beta}$ through the instrument moments
$\mathbf{W}'(\mathbf{y} - \mathbf{X}\boldsymbol{\beta} - \mathbf{e}) = \mathbf{0}$,
where $\mathbf{W}$ is the instrument matrix [@golan2008information, pp. 89-91],
and keeps the same support-and-entropy machinery. It uses a matrix interface,
`linreg_iv(y, X, IV, Z)`, and supports both just-identified ($\text{ncol}(IV) =
\text{ncol}(X)$) and over-identified systems.

We simulate an endogenous regressor $x_e$ that shares a shock $u$ with the error,
and an instrument $z$ that drives $x_e$ but is independent of $u$:

```{r iv-dgp}
set.seed(125)
n  <- 200
z  <- rnorm(n)                       # instrument
u  <- rnorm(n)                       # shared shock => endogeneity
xe <- 0.7 * z + u + rnorm(n)         # endogenous regressor
y  <- 1 + 1.5 * xe + u               # true slope = 1.5

X   <- cbind(`(Intercept)` = 1, xe = xe)   # design: intercept + endogenous xe
IV  <- cbind(`(Intercept)` = 1, z  = z)    # instruments: intercept (self) + z
Ziv <- matrix(c(-10, 0, 10), nrow = 2, ncol = 3, byrow = TRUE)  # signal support
```

Because $x_e$ is positively correlated with $u$ and $u$ also enters $y$, OLS
over-states the slope. Two-stage least squares (2SLS) and GME-IV both use $z$ to
purge the endogeneity.

```{r iv-fit}
ols_iv <- lm(y ~ xe)
b_2sls <- solve(crossprod(IV, X), crossprod(IV, y))   # just-identified 2SLS
gme_iv <- linreg_iv(y, X, IV, Ziv)

tab_iv <- data.frame(
  Truth    = c(`(Intercept)` = 1, xe = 1.5),
  OLS      = coef(ols_iv),
  `2SLS`   = as.vector(b_2sls),
  `GME-IV` = coef(gme_iv),
  check.names = FALSE
)
knitr::kable(tab_iv, digits = 3,
             caption = "OLS is biased by endogeneity; 2SLS and GME-IV recover the true slope.")
```

OLS pulls the slope above its true value of 1.5, while both instrumented
estimators bring it back close to the truth. As the signal support widens the
GME-IV estimate approaches the exact 2SLS solution; a narrower support
regularizes it, which can help when the instruments are weak. `summary(gme_iv)`
reports the instrument-moment residual (near zero at convergence) and the
normalized signal entropy.

# Practical guidance

- **Supports are the key input.** Center the coefficient support on plausible
  values and make it wide enough to contain the truth; a wider support behaves
  more like OLS, a narrower one regularizes more strongly. Use the three-sigma
  rule for the symmetric noise support `v`.
- **Reach for GCE when you have real prior information.** Encode it through
  `p0`; a uniform `p0` is exactly GME.
- **Leave `nu = 0.5` unless you have a reason not to.** It is the standard
  equal-weight GME.
- **GME shines under collinearity and other ill-conditioned designs**, where it
  trades a little bias for a large reduction in variance and instability.
- **Report the entropy diagnostics.** `summary()` gives the entropy-ratio test
  and the normalized entropy $S$; the pseudo-$R^2 = 1 - S$ and the Wald test
  (Section \@ref(inference)) complete the inferential toolkit.
- **Endogeneity? Reach for `linreg_iv()`.** Supply an instrument matrix (each
  exogenous regressor instruments itself); it accepts just- and over-identified
  systems and shares `linreg()`'s supports and priors.

See `?linreg` and `?linreg_iv` for the full argument lists, `summary()` for the
entropy-ratio coefficient test, and the *Pure Inverse Problems* vignette for the
underlying maximum-entropy machinery.

# References
