---
title: "Introduction to infometrics"
author: "Package Author"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to infometrics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6,
  fig.height = 4
)
```

## Overview

The **infometrics** package implements the class of Information-Theoretic (IT)
estimators for econometric models, following the unified framework developed
in Golan (2008). The connecting thread across all estimators is the
**Cressie-Read criterion** — a generalized entropy of order α — of which
Shannon entropy, Empirical Likelihood, and the chi-squared statistic are all
special cases.

Phase 1 provides:

- Entropy and divergence measures (Shannon, KL, Rényi, Tsallis, Cressie-Read)
- The classical Maximum Entropy (ME) and Cross-Entropy (CE) estimators
- Inference tools (normalized entropy, entropy-ratio statistic, pseudo-R²)
- Support space constructors for GME/GCE (Phase 2)

```{r load}
library(infometrics)
```

---

## 1. Entropy measures

### Shannon entropy

Shannon entropy H(p) = -Σ p_k log(p_k) measures the uncertainty in a
probability distribution. It reaches its maximum value of log(K) for the
uniform distribution over K outcomes (complete ignorance) and equals zero for
a degenerate point mass (perfect certainty).

```{r shannon}
# Uniform distribution over 6 outcomes: maximum entropy
p_uniform <- rep(1/6, 6)
shannon_entropy(p_uniform)       # = log(6)
log(6)

# A more informative distribution: lower entropy
p_skewed <- c(0.50, 0.25, 0.10, 0.08, 0.05, 0.02)
shannon_entropy(p_skewed)

# In bits (base 2): H(0.5, 0.5) = 1 bit
shannon_entropy(c(0.5, 0.5), base = 2)
```

### The Cressie-Read unifying family

The `cressie_read()` function implements the power divergence statistic that
connects all IT estimators through a single parameter α (Golan 2008, Eq. 3.8).

```{r cr-family}
p <- c(0.50, 0.30, 0.20)
q <- rep(1/3, 3)

# alpha -> 0: KL divergence
cat("CR(alpha=1e-9):   ", cressie_read(p, q, alpha = 1e-9), "\n")
cat("KL divergence:    ", kl_divergence(p, q), "\n")

# alpha = 1: Pearson chi-squared / 2
cat("CR(alpha=1):      ", cressie_read(p, q, alpha = 1), "\n")
cat("Pearson chisq/2:  ", 0.5 * sum((p - q)^2 / q), "\n")
```

### Normalized entropy

The normalized entropy S(p̃) = H(p) / log(K) is the primary goodness-of-fit
measure for IT models. It lies in [0, 1], with 1 indicating complete ignorance
and 0 indicating perfect certainty.

```{r normalized}
normalized_entropy(p_uniform)  # = 1
normalized_entropy(p_skewed)   # < 1
```

---

## 2. Maximum Entropy estimation

### 2.1 Under the Hood: the Dual ME by Hand

Before calling `me()`, it is instructive to build the dual ME estimator from
scratch for the scalar-constraint case. Jaynes (1957) posed the following
problem: a die is thrown many times and only the mean outcome ȳ = 4.5 is
recorded. What is the maximum-entropy distribution consistent with that mean?

The **dual concentrated problem** (Golan 2008, Eq. 4.4–4.5) replaces the
K-dimensional probability vector **p** with a single Lagrange multiplier λ:

```
min_λ  −λ ȳ + log Ω(λ),    Ω(λ) = Σ_k q_k exp(λ x_k)
```

where **q** is the prior (uniform by default) and x_k are the support points.
The solution is p̂_k = q_k exp(λ̂ x_k) / Ω(λ̂).

The implementation below uses the **log-sum-exp trick** for numerical stability
and **BFGS with an analytic gradient** — the same choices made inside `me()`.

```{r jaynes-helpers}
# Internal helpers: dot-prefix = not exported
.jd_obj <- function(lambda, x, q, y) {
  e         <- lambda * x               # exponents
  m         <- max(e)                   # log-sum-exp shift
  log_Omega <- log(sum(q * exp(e - m))) + m
  -lambda * y + log_Omega               # dual objective (Golan 2008, Eq. 4.4)
}

.jd_grad <- function(lambda, x, q, y) {
  e   <- lambda * x
  m   <- max(e)
  w   <- q * exp(e - m)
  p   <- w / sum(w)                     # p_k = q_k exp(lambda x_k) / Omega
  -(y - sum(x * p))                     # gradient (Golan 2008, Eq. 4.5)
}

jaynes_est <- function(x, y, q = NULL) {
  K <- length(x)
  if (K < 2)
    stop("x must have length >= 2. Got length(x) = ", K, ".")
  if (y <= min(x) || y >= max(x))
    stop("y must satisfy min(x) < y < max(x). Got y = ", y,
         ", min(x) = ", min(x), ", max(x) = ", max(x), ".")
  if (is.null(q))
    q <- rep(1 / K, K)
  if (length(q) != K || any(q <= 0) || abs(sum(q) - 1) > 1e-8)
    stop("q must be a probability vector of length ", K, ".")

  opt    <- stats::optim(0, .jd_obj, .jd_grad, x = x, q = q, y = y,
                         method = "BFGS")
  lambda <- opt$par
  e      <- lambda * x
  m      <- max(e)
  w      <- q * exp(e - m)
  p_hat  <- w / sum(w)

  list(
    p_hat           = p_hat,
    lambda_hat      = lambda,
    H_star          = -sum(p_hat * log(p_hat)),
    moment_residual = y - sum(x * p_hat),
    converged       = (opt$convergence == 0)
  )
}
```

```{r jaynes-demo}
x      <- 1:6
y_mean <- 4.5

jfit <- jaynes_est(x, y_mean)
round(jfit$p_hat, 4)
cat("lambda_hat:     ", round(jfit$lambda_hat, 6), "\n")
cat("H_star:         ", round(jfit$H_star, 6),     "\n")
cat("moment_residual:", format(jfit$moment_residual, scientific = TRUE), "\n")
cat("converged:      ", jfit$converged,             "\n")
```

The moment residual is effectively zero, confirming the constraint
ȳ = Σ x_k p̂_k is satisfied. We verify that `me()` produces identical
probabilities:

```{r jaynes-equiv}
fit_me_check <- me(y = 4.5, X = matrix(1:6, nrow = 1))
cat("max |p_hat_manual - p_hat_me|:",
    max(abs(jfit$p_hat - coef(fit_me_check))), "\n")
```

---

### 2.2 The classic "loaded die" example via `me()`

Following Golan (2008, Section 4.6): we observe only the mean of a six-sided
die throw (y = 4.5) and want to recover the underlying probability
distribution. This is the canonical under-determined problem (K = 6
unknowns, T = 1 constraint).

```{r die-problem}
support <- 1:6
X       <- matrix(support, nrow = 1)  # 1 moment, 6 outcomes
y_mean  <- 4.5                        # observed mean

fit <- me(y = y_mean, X = X)
print(fit)
```

The estimated probabilities show more mass concentrated on higher outcomes, as
required by the mean constraint. The normalized entropy S is close to 1,
meaning the single mean constraint leaves substantial uncertainty.

```{r die-plot, fig.cap="ME estimated probabilities for the die problem"}
barplot(
  fit$p_hat,
  names.arg = 1:6,
  xlab = "Outcome",
  ylab = "Probability",
  main = paste0("ME estimates  (S = ", round(fit$S, 3), ")"),
  col  = "steelblue"
)
abline(h = 1/6, lty = 2, col = "red")
legend("topleft", legend = "Uniform (1/6)", lty = 2, col = "red", bty = "n")
```

### Adding a second moment constraint

Adding E[X²] = 25 (higher variance) further constrains the distribution:

```{r die-two-moments}
X2 <- rbind(1:6, (1:6)^2)
y2 <- c(4.5, 25)

fit2 <- me(y = y2, X = X2)
summary(fit2)
```

With two constraints, S drops — the data are more informative. Both moment
constraints are satisfied up to numerical precision.

### Cross-Entropy with a non-uniform prior

When prior information is available, the CE formulation incorporates it by
minimizing D(p‖q) relative to the prior q rather than the uniform
distribution.

```{r ce}
# Prior: belief that the die is biased toward high numbers
q_prior <- c(0.05, 0.08, 0.12, 0.18, 0.27, 0.30)
sum(q_prior)  # must sum to 1

fit_ce <- me(y = 4.5, X = matrix(1:6, nrow = 1), q = q_prior)

# Compare ME (uniform prior) vs CE (informative prior)
rbind(
  ME = round(fit$p_hat, 4),
  CE = round(fit_ce$p_hat, 4)
)
```

The CE solution shifts probability mass toward outcomes favored by the prior,
relative to the ME solution.

---

## 3. Support space construction

Support spaces bound the parameter and error vectors in GME/GCE estimation
(Phase 2). The `make_support()` function creates symmetric M-point grids.

```{r supports}
# 5-point symmetric support for an error term: [-3, -1.5, 0, 1.5, 3]
make_support(half_range = 3, M = 5)

# Data-driven supports from OLS (for use in GME)
set.seed(42)
n <- 100; K <- 3
Xreg <- cbind(1, matrix(rnorm(n * (K - 1)), n, K - 1))
yreg <- Xreg %*% c(1, 2, -1) + rnorm(n)

sp <- default_supports(yreg, Xreg)
cat("Signal support Z (row = coefficient):\n")
print(round(sp$Z, 3))
cat("\nError support V:\n")
print(round(sp$V, 3))
```

The default signal supports are centered on OLS estimates with half-range
equal to 2×|OLS estimate|. The error support uses the three-sigma rule
(Pukelsheim 1994), spanning ±3 OLS residual standard deviations.

---

## 4. Inference

The `summary()` method reports the full inference output including the
entropy-ratio test statistic W, which under H₀ (all parameters = 0) converges
to χ²(K-1):

```{r inference}
# Loaded die with strong evidence (mean = 5.5, far from uniform)
fit_strong <- me(y = 5.5, X = matrix(1:6, nrow = 1))
summary(fit_strong)
```

A high W statistic and low p-value indicate the data strongly reject the
null of a uniform distribution, pulling the estimates away from complete
ignorance.

---

## References

Golan, A. (2008). Information and Entropy Econometrics — A Review and
Synthesis. *Foundations and Trends in Econometrics*, **2**(1–2), 1–145.

Jaynes, E.T. (1957). Information theory and statistical mechanics.
*Physical Review*, **106**, 620–630.

Cressie, N. and Read, T.R.C. (1984). Multinomial goodness-of-fit tests.
*Journal of the Royal Statistical Society B*, **46**, 440–464.

Kullback, S. and Leibler, R.A. (1951). On information and sufficiency.
*Annals of Mathematical Statistics*, **22**, 79–86.

Pukelsheim, F. (1994). The three sigma rule.
*The American Statistician*, **48**, 88–91.
